---
output: html_document
editor_options: 
  chunk_output_type: console
---
% Managing Data with `dplyr` and `tidyr`
% Biostatistics 140.776
%

```{r, include=FALSE}
options(width = 50)
knitr::opts_chunk$set(comment = NA)
```

# dplyr

The data frame is a key data structure in statistics and in R.

* There is one observation per row

* Each column represents a variable or measure or characteristic

* Primary implementation that you will use is the default R
  implementation

* Other implementations, particularly relational databases systems


# dplyr

* Developed by RStudio

* An optimized and distilled version of `plyr` package (by Hadley Wickham)

* Provides a "grammar" (in particular, verbs) for data manipulation

* Is **very** fast, as many key operations are coded in C++

* Functions (verbs) can be chained together via a "pipe" operator


# dplyr Verbs

* `select`: return a subset of the columns of a data frame

* `filter`: extract a subset of rows from a data frame based on
  logical conditions

* `arrange`: reorder rows of a data frame


* `rename`: rename variables in a data frame

* `mutate`: add new variables/columns or transform existing variables

* `summarize` / `summarise`: generate summary statistics of different variables in the data frame, possibly within strata

There is also a handy `print` method that prevents you from printing a lot of data to the console.



# dplyr Properties

* The first argument is a data frame.

* The subsequent arguments describe what to do with it, and you can
  refer to columns in the data frame directly without using the $
  operator (just use the names).

* The result is a new data frame

* Data frames must be properly formatted and annotated for this to all
  be useful
  
* There are no "inputs" and "results"; it's all just *data*


# Load the `dplyr` package


This step is important!

```{r}
library(dplyr)
```


# `select`

```{r,message=FALSE}
library(readr)
specdata <- read_csv("SPEC_2014.csv.gz")
dim(specdata)
head(select(specdata, 1:5))
```


# `select`

```{r}
names(specdata)[1:3]
head(select(specdata, State.Code:Site.Num))
```

# `select`

In dplyr you can do

```{r,eval=FALSE}
head(select(specdata, -(State.Code:Site.Num)))
```

Equivalent base R

```{r,eval=FALSE}
i <- match("State.Code", names(specdata))
j <- match("Site.Num", names(specdata))
head(specdata[, -(i:j)])
```



# `filter`

```{r}
specdata.f <- filter(specdata, Parameter.Name == "Sulfate PM2.5 LC")
select(specdata.f, 1:3, Parameter.Name)
```

# `filter`

```{r}
specdata.f <- filter(specdata, Parameter.Name == "Sulfate PM2.5 LC" 
                     & State.Name == "Texas")
select(specdata.f, 1:3, Parameter.Name, State.Name)
```


# `arrange`

Reordering rows of a data frame (while preserving corresponding order
of other columns) is normally a pain to do in R.

```{r}
specdata <- arrange(specdata, Date.Local)
head(select(specdata, Date.Local, Parameter.Name, Sample.Value), 3)
tail(select(specdata, Date.Local, Parameter.Name, Sample.Value), 3)
```

# `arrange`

Columns can be arranged in descending order too.

```{r}
specdata <- arrange(specdata, desc(Date.Local))
head(select(specdata, Date.Local, Parameter.Name, Sample.Value), 3)
tail(select(specdata, Date.Local, Parameter.Name, Sample.Value), 3)
```


# `rename`

Renaming a variable in a data frame in R is surprising hard to do!

```{r,tidy=FALSE}
head(specdata[, 1:7], 3)
specdata <- rename(specdata, lat = Latitude, 
                   lon = Longitude)
head(specdata[, 1:7], 3)
```


# `mutate`

```{r, tidy=FALSE}
specdata <- mutate(specdata, 
                   city_state = paste(City.Name, State.Name, sep = ", "),
                   sample_mg = Sample.Value / 1000)
head(select(specdata, City.Name, State.Name, city_state, sample_mg))
```

# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
specdata <- mutate(specdata, 
                   region = factor(lon > -100, 
                                   labels = c("west", "east")))
eastwest <- group_by(specdata, region)
summarize(eastwest, pollutant = mean(Sample.Value, na.rm = TRUE),
          obs = mean(Observation.Count, na.rm = TRUE))
```


# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
library(lubridate)
specdata <- mutate(specdata, month = month(Date.Local))
months <- group_by(specdata, month)
months <- filter(months, Parameter.Name == "Sulfate PM2.5 LC")
summarize(months, sulfate = mean(Sample.Value, na.rm = TRUE))
```

```{r,echo=FALSE}
specdata$month <- NULL  ## Can't use mutate to create an existing variable
```


# `%>%` (pipe operator)

```{r,tidy=FALSE,eval=FALSE}
specdata %>% 
        mutate(month = month(Date.Local)) %>% 
        filter(Parameter.Name == "Sulfate PM2.5 LC") %>%
        group_by(month) %>% 
        summarize(sulfate = mean(Sample.Value, na.rm = TRUE))
```

```{r,echo=FALSE}
specdata %>% 
        mutate(month = month(Date.Local)) %>% 
        filter(Parameter.Name == "Sulfate PM2.5 LC") %>%
        group_by(month) %>% 
        summarize(sulfate = mean(Sample.Value, na.rm = TRUE))
```

# `%>%` (pipe operator)

```{r,tidy=FALSE,eval=FALSE}
specdata %>% 
        mutate(month = month(Date.Local)) %>% 
        filter(Parameter.Name == "Sulfate PM2.5 LC") %>%
        group_by(month, region) %>% 
        summarize(sulfate = mean(Sample.Value, na.rm = TRUE))
```

```{r,echo=FALSE}
specdata %>% 
        mutate(month = month(Date.Local)) %>% 
        filter(Parameter.Name == "Sulfate PM2.5 LC") %>%
        group_by(month, region) %>% 
        summarize(sulfate = mean(Sample.Value, na.rm = TRUE))
```


# dplyr

Once you learn the dplyr "grammar" there are a few additional benefits

* dplyr can work with other data frame "backends"

* `data.table` for large fast tables

* SQL interface for relational databases via the DBI package

# tidyr

The `tidyr` package helps with manipulation of data frames between "wide" and "long" formats, depending on what you're trying to do.

* Sometimes the meaning of a "variable" depends on the application

* Sometimes PM10, O3, NO2 are all different variables with continuous levels

* Sometimes "Pollutant" is the variable with levels "PM10", "O3", and "NO2"

# Long Format

```{r, include=FALSE}
spec <- select(specdata, city_state, Date.Local, Parameter.Name, Sample.Value)
```

Here are the specdata pollution data in long format

```{r}
head(spec)
```

# Long Format

```{r}
stats <- group_by(spec, Parameter.Name) %>% 
        summarize(mean = mean(Sample.Value, na.rm = TRUE),
                  median = median(Sample.Value, na.rm = TRUE),
                  max = max(Sample.Value, na.rm = TRUE))
stats
```

# `gather`

An alternate representation could have three variables: pollutant, statistic, and value

* `gather` is a function that "gathers" multiple columns and essential sticks them into one column

* The names of multiple columns become levels of a single variable

* In this case mean, median, max --> levels of a "statistic" variable

# `gather`

```{r}
library(tidyr)
gather(stats, statistic, value, -Parameter.Name)
```

# `spread`

The `spread` function does the inverse of the `gather` function

* `spread` takes a single variable (with multiple levels) and *spreads* them across multiple columns

* Sometimes more intuitive if you want to compute a statistic across multiple levels/variables

* e.g. Compute the maximum of three different pollutants on each day and create a new variable

# `spread`

```{r}
wide <- filter(spec, city_state == "Essex, Maryland") %>%
        spread(Parameter.Name, Sample.Value) 
head(wide)
```

# `separate`

Sometimes you need to split one column into two separate columns.

```{r, include=FALSE}
x <- paste(rep(c("happy", "sad"), each = 3), rep(c("working", "partying"), 3), sep = "-")
m <- data.frame(mood = x, day = c(rep("Sunday", 4), rep("Monday", 2)))
```

```{r}
m
```

# `separate`

```{r}
m <- separate(m, mood, c("state", "activity"), sep = "-")
m
```

# `unite`

The inverse of `separate`

```{r}
unite(m, mood, state, activity, sep = "-")
```

# Summary

`dplyr`

* Verbs/functions for manipulating data frames in tidy format

* select, filter, arrange, group_by, summarize, rename, mutate

`tidyr`

* Transform data frames from wide to long formats

* spread, gather, separate, unite
