---
title: "Joining Data"
subtitle: Biostatistics 140.776
output:
  beamer_presentation: default
  ioslides_presentation: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = NA)
```



## Joining Data

The `dplyr` package provides a set of functions for joining two data frames into a single data frame based on a set of key columns.

* `left_join()`

* `right_join()`

* `inner_join()`

There are other functions for joining but they are less commonly used.


## Left Join

```{r, message = FALSE}
library(dplyr)


dat <- tibble(
        id = rep(c("a", "b", "c"), each = 3),
        visit = rep(0:2, 3),
        outcome = rnorm(3 * 3, 3)
)
```

## Left Join

```{r}
dat
```

## Left Join

```{r}
subjects <- tibble(
        id = c("a", "b", "c"),
        house = c("detached", "rowhouse", "rowhouse"),
)
```
```{r}
subjects
```


## Left Join

```{r}
left_join(dat, subjects, by = "id")
```

## Left Join

```{r}
subjects <- tibble(
        id = c("a", "b", "c"),
        visit = c(0, 1, 0),
        house = c("detached", "rowhouse", "rowhouse"),
)
```
```{r}
subjects
```

## Left Join

```{r}
left_join(dat, subjects, by = c("id", "visit"))
```


## Left Join

```{r}
subjects <- tibble(
        id = c("b", "c"),
        visit = c(1, 0),
        house = c("rowhouse", "rowhouse"),
)
```
```{r}
subjects
```

## Left Join

```{r}
left_join(dat, subjects, by = c("id", "visit"))
```


## Inner Join


```{r}
inner_join(dat, subjects, by = c("id", "visit"))
```



## Right Join

```{r}
right_join(dat, subjects, by = c("id", "visit"))
```



## Right Join

```{r}
right_join(subjects, dat, by = c("id", "visit"))
```


## Summary

* `left_join()` is useful for merging a "large" data frame with a "smaller" one while retaining all the rows of the "large" data frame

* `inner_join()` gives you the intersection of the rows between two data frames 

* `right_join()` is like `left_join()` with the arguments reversed (likely only useful at the end of a pipeline)


